Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-07-06 unimodular-covers.tex
reviewed introduction

\paco{define mixed IDP, Cayley sums, Minkowski sums, mixed subdivisions?}
A lattice polytope is called \emph{smooth} if it is simple and the edge generators at every vertex form a linear basis for the lattice. 
%
Oda~\cite{Oda1997} posed several questions regarding IDP polytopes which are now considered conjectures~\cite{HNPS2008,mfo2007}. 
Among them:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item (Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Let $P, Q\subset \R^d$ be lattice polytopes with the normal fan of $Q$ refining that of $P$, and $P$ smooth. Then
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
\paco{perhaps we can call this formula ``being mixed IDP''. The property of ``$N(Q)$ refines $N(P)$'' is sometimes called ``$P$ is a Minkowski summand of $Q$''.}

@@

This not only implies \Cref{conj:Oda}(1) in dimension three, but also \Cref{conj:Oda}(2) for $3$-dimensional \emph{prismatoids}, polytopes whose vertices all lie in two parallel facets:

@@

%\section{Preliminaries}
%\label{sec:prelim}

@@

Let $w=\area(p+q) \geq 2$, where $\area$ denotes the area normalized to a fundamental domain. Then 
\[
w=\width_{f_q}(p + \vecline q )=\width_{f_q}(p)=\width_{f_p}(q)=\width_{f_p}(q +\vecline p ).
\]

Suppose by contradiction that there is no lattice point as described in the lemma. In particular, no vertex of $P$ can be in the strip $p + \vecline q$.  Thus $P$ has two primitive segments in the boundary which each have one vertex on each side of the strip $p + \vecline q$; we will call these $\ell=[l_1, l_2]$ and $r=[r_1, r_2]$, with $\ell$ and $r$ crossing the strip $p + \vecline q$  in $V_1$ and $ V_2$ respectively and the convention that $f_q(l_2) >f_q(l_1)$ and $f_q(r_2) >f_q(r_1)$. This readily implies 
\begin{gather}
\label{eq:widthp}
\begin{array}{cc}
f_q(l_2) \geq f_q(p_2), &
f_q(l_1) \leq f_q(p_1), \\
f_q(r_2) \geq f_q(p_2), &
f_q(r_1) \leq f_q(p_1).
\end{array}
\end{gather}

@@

\begin{claim}
$f_q(b_2-b_1)$ and $f_q(p_2 - p_1)$ are non-zero and have the same sign. That is, $f_q$ achieves its maximum over $b$ and over $t$ on the same halfplane $V_1$ or $V_2$.
\end{claim}\giulia{do we need this claim?}

@@

The last two claims can be summarized as saying that in the pictures $b$, $t$ and $r$ have positive slope.

@@

\begin{proof}[Proof of \Cref{lemma:cayley}]
We are now ready to show a contradiction. Since the normal fan of $Q$ refines that of $P$, $Q$ must have an edge $r'$ which is a translated copy of $r$. Let $r_1'$ and $r_2'$ be its endpoints. Now consider the lattice line $d$ through $r_1'$ parallel to  $[p_1, q_2]$, that is, $g$ is constant on $d$. Let $d'$ be the parallel line defined by $g(d')=g(d)+1$.

>>>>

\paco{define mixed IDP, Cayley sums, Minkowski sums, mixed subdivisions?}
A lattice polytope is called \emph{smooth} if it is simple and the edge generators at every vertex form a linear basis for the lattice. 
%
Oda~\cite{Oda1997} posed several questions regarding IDP polytopes. The following variations of them are now considered conjectures~\cite{HNPS2008,mfo2007}:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item (Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Let $P, Q\subset \R^d$ be lattice polytopes with the normal fan of $Q$ refining that of $P$, and $P$ smooth. Then
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
\paco{perhaps we can call this formula ``being mixed IDP''. The property of ``$N(Q)$ refines $N(P)$'' is sometimes called ``$P$ is a Minkowski summand of $Q$''.}

@@

This not only implies \Cref{conj:Oda}(1) in dimension three, but also, as we show in \Cref{sec:cayley}, \Cref{conj:Oda}(2) for $3$-dimensional \emph{prismatoids}, polytopes whose vertices all lie in two parallel facets:

@@

%\section{Preliminaries}
%\label{Fprelim}

@@

Let $w=\area(p+q) \geq 2$, where $\area$ denotes the area normalized to a fundamental domain. Then 
\[
w=\width_{f_q}(p + \vecline q )=\width_{f_q}(p)=\width_{f_p}(q)=\width_{f_p}(q +\vecline p ). 
\] \giulia{we want a picture for the setup}

\begin{proof}[Proof of \Cref{lemma:cayley}]
Suppose by contradiction that there is no lattice point as described in the lemma. In particular, no vertex of $P$ can be in the strip $p + \vecline q$.  Thus $P$ has two primitive segments in the boundary which each have one vertex on each side of the strip $p + \vecline q$; we will call these $\ell=[l_1, l_2]$ and $r=[r_1, r_2]$, with $\ell$ and $r$ crossing the strip $p + \vecline q$  in $V_1$ and $ V_2$ respectively and the convention that $f_q(l_2) >f_q(l_1)$ and $f_q(r_2) >f_q(r_1)$. This readily implies 
\begin{gather}
\label{eq:widthp}
\begin{array}{cc}
f_q(l_2) \geq f_q(p_2), &
f_q(l_1) \leq f_q(p_1), \\
f_q(r_2) \geq f_q(p_2), &
f_q(r_1) \leq f_q(p_1).
\end{array}
\end{gather}

@@

\begin{claim}
$f_q(b_2-b_1)$ and $f_q(t_2 - t_1)$ are non-zero and have the same sign. That is, $f_q$ achieves its maximum over $b$ and over $t$ on the same halfplane $V_1$ or $V_2$.
\end{claim}

@@

The last two claims can be summarized as saying that in the pictures $b$, $t$ and $r$ have positive slope. Observe that this implies that $q$ is not in the boundary of $Q$ and $p \neq r$, so both $P$ and $Q$ are full dimensional.

@@

We are now ready to show a contradiction. Since the normal fan of $Q$ refines that of $P$, $Q$ must have an edge $r'$ which is a translated copy of $r$. Let $r_1'$ and $r_2'$ be its endpoints. Now consider the lattice line $d$ through $r_1'$ parallel to  $[p_1, q_2]$, that is, $g$ is constant on $d$. Let $d'$ be the parallel line defined by $g(d')=g(d)+1$.
